Information in recombination junctions can both compress and improve inferred tree sequences

### 2024-01-04 paper/writeup.tex
added figures, extend edges section

\newcommand{\E}{\mathbb{E}}
\renewcommand{\P}{\mathbb{P}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\tn}{\textnormal}
\newcommand{\ov}{\overline}
\newcommand{\tskit}{\texttt{tskit}}
\newcommand{\comment}[1]{{\color{blue} \it #1}}

@@

\begin{figure}
    \begin{center}
    \includegraphics[width=0.9\textwidth]{conceptual_figure}
    \end{center}
    \caption{
        A simple example showing the basic idea
        (described in more detail in the text):
        \textbf{(A)} a small portion of a tree sequence without unary nodes;
        \textbf{(B)} the implied inheritance pattern of the two portions of the haplotype carried by ancestral node 4,
        labeled $L$ and $R$;
        \textbf{(C)} marginal trees with a unary node added,
        which produces \textbf{(D)} a more parsimonious haplotype inheritance pattern
        (that also includes fewer edges).
        \label{fig:conceptual}
    }
\end{figure}

@@

% * description (HALLEY)
% * proof of something:
%     - arrives at a local minimum?
%     - guess at typical reduction? (reduces edges by 1/3?)
    We define a \textit{tree sequence}, denoted $TS$, 
    as a tuple $\left( \left\{T_k\right\}, N, E, L\right)$ where
     $\left\{T_k\right\}_{k=1}^n$ is the set of trees,
     $N$ and $E$ are the sets of nodes and edges respectively,
     and $L=[0,a)\subseteq \R$ is the length of the sequence on the genome.
     The length of each tree $T_k$ is some subinterval $[a_k,a_{k+1})$ 
     with the lengths of consecutive trees in the sequence 
     only intersecting each intervals endpoints.
    Given any tree sequence, our goal is to
    identify areas of implied inheritance of haplotypes.
    We do so by identifying short paths (2 edges) in a tree.
    If neighboring trees contain edges with 
    identical parent and child nodes, we assume that there
    is implied coalescence of haplotypes between nodes, and
  	we will extend the path's edges to the neighboring tree \ref{fig:extending_diagram}.
  	In more concrete terms, for a tree $T_k$ in the tree sequence $TS$ 
    suppose there exists a path which contains a node $b\in N$,
    with parent and child nodes $a, c\in N$, respectively.
    If there is an edge between the nodes $a$ and $c$ in trees 
    $T_{k+1}$ (or $T_{k-1}$),
    then we wish to extend the edges $e_{ab},e_{bc}$ 
    to $T_{k+1}$ and then remove the edge $e_{ac}$ from $T_{k+1}$. 
    This action reduces the length of $e_{ac}$ on the genome,
    and in some cases, completely removes the edge. 
    We now perform this action on all such ancestral paths 
    over the entire tree sequence.

\begin{figure}[!ht]
\begin{center}
\begin{tikzpicture} 
	\node[circle,draw,inner sep=1pt] (a) at (0,2) {a}
	[growth parent anchor=south] 
		child { node[circle,draw,inner sep=1pt] {b} 		
	 			child {
	 				node[circle,draw,inner sep=1pt] (c) at (-0.5,0) {c}
	 				edge from parent 
	 					node[left] {$e_{b,c}$}}
 				child {
 					node (d) at (0,-1) {$T_k$}
 					edge from parent [draw=none]}
	 			child {
	 				node[circle,draw,inner sep=1pt] (e) at (0.5,0)  {d}
 					edge from parent
 						node[right] {$e_{b,d}$}} 		
	 		edge from parent
	 			node[left] {$e_{a,b}$}
	 	}; 	
	\node[circle,draw,inner sep=1pt] (a') at (4,2) {a}
	[growth parent anchor=south]
		child {
			child{
				node[circle,draw,inner sep=1pt] (c') at (0,-0.25) {c}	
					child{node (d) at (0,1) {$T_{k+1}$}
						edge from parent [draw=none]}
			}
			node[left] {$e_{a,c}$}
		};

	\node at (7,0) {$\Longrightarrow$};
	\node at (7,0.5) {\tn{Edge Extend}};

	\node[circle,draw,inner sep=1pt] (a''') at (9,2) {a}
	[growth parent anchor=south] 
	child { node[circle,draw,inner sep=1pt]	 {b}	
		child {
			node[circle,draw,inner sep=1pt] (c) at (0,0) {c}
			child{node (d) at (0,1) {$T_k$}
				edge from parent [draw=none]}
			edge from parent 
			node[left] {$e_{b,c}$}
		}
		edge from parent
		node[left] {$e_{a,b}$}
	};

	\node[circle,draw,inner sep=1pt] (a'v) at (12,2) {a}
	[growth parent anchor=south] 
	child { node[circle,draw,inner sep=1pt] {b}		
		child {
			node[circle,draw,inner sep=1pt] (c) at (0,0) {c}
			child{node (d) at (0,1) {$T_{k+1}$}
				edge from parent [draw=none]}
			edge from parent 
			node[left] {$e_{b,c}$}
		}
		edge from parent
		node[left] {$e_{a,b}$}
	};

\end{tikzpicture}
\end{center}
\caption{A simple visualization of the Edge Extend method.
    \label{fig:extending_diagram}
}
\end{figure}

    In more detail, let $TS = \left( \left\{T_k\right\}_{k=1}^n, N, E, L\right)$ 
    be a tree sequence.
    Within tskit we can track a tree sequence using a function called `EdgeDiffs'.
    This function tracks edges which are removed and added 
    between consecutive trees in the sequence. 
    As in \ref{fig:extending_diagram} (tikzfigure), we can track if 
    $e_{a,b}$ and$e_{b,c}$ were removed from $T_k$ 
    when creating $T_{k+1}$, 
    and we can see if $e_{a,c}$ was just added in $T_{k+1}$.

@@

\caption{Extends edges across a tree sequence $TS$.}\label{alg:edge}
\SetKwFunction{ex}{Extend Edges}
\SetKwFunction{fwd}{Forward Extend}
\DontPrintSemicolon
\LinesNotNumbered
%\SetAlgoNoEnd
\input{A tree sequence $TS$}
\output{New tree sequence,$TS'$, with reduced number of edges}

@@

\Fn{\fwd (tree sequence $TS$, forwards = \tn{True})}{
{	
\eIf{forwards=\tn{True}}{EdgeDiffs $\gets TS$.edgediffs()\BlankLine}
{\BlankLine
	EdgeDiffs $\gets$ reversed($TS$.edgediffs())\;}
}
\Indm \For{(Interval, EdgesOut, EdgesIn) \KwTo EdgeDiffs}{
	\For{e1 \KwTo EdgesOut}{
		\For{e2 \KwTo EdgesOut}{
			\If{e2.parent = e1.child}{
				\For{e3 \KwTo EdgesIn}{
					\If{e3.parent=e1.parent \tn{AND} e3.child=e2.child}{
						e1.right $\gets$ Interval.right\;
						e2.right $\gets$ Interval.right\;
						e3.left $\gets$ Interval.right\;
						\If{e3.left = e3.right}{
							$E$.remove(e3)\;
							}}}}}}}\;
$TS'\gets TS$.update()\;						
\KwRet{$TS'$}\;
}
\BlankLine						
\Fn{\ex ($TS$, iter=100)}{
	edges $\gets$ $TS$.numedges()\;
	\BlankLine
	\For{$*$ \KwTo \Range(iter)}{
		$TS' \gets TS$.\fwd(fowards=\tn{True})\;
		$TS' \gets TS$.\fwd(fowards=\tn{False})\;
		\eIf{$TS$.numedges = edges}{
			\Break\BlankLine}
		{edges $\gets$ $TS'$.numedges()\;}
		}
}
\KwRet{$TS'$}\;
\end{algorithm2e}

With this algorithm we only extend existing edges
over a larger interval on the genome,
and remove unnecessary edges in the process.
%% Find counter example to global minimum.

@@

\par Now with an algorithm in place
 we need a metric to discern
 if trees modified with edge extend
 in fact correctly infer haplotypes in our original tree sequence.
 We have thus constructed a function which counts the differences 
 between trees on a pair of tree sequences. 
 To count all the discrepancies we must begin by counting differences for each node in the tree sequence.

\begin{figure}
    \caption{
        Conceptual figure displaying what discrepancy is?
        TODO: brainstorm.
        \label{fig:conceptual_discrepancy}
    }
\end{figure}

>>>>

\newcommand{\E}{\mathbb{E}}
\renewcommand{\P}{\mathbb{P}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\T}{\mathbb{T}}
\newcommand{\tn}{\textnormal}
\newcommand{\ov}{\overline}
\newcommand{\tskit}{\texttt{tskit}}
\newcommand{\comment}[1]{{\color{blue} \it #1}}

@@

\begin{figure}
    \begin{center}
    \includegraphics[width=0.9\textwidth]{conceptual_figure.pdf}
    \end{center}
    \caption{
        A simple example showing the basic idea
        (described in more detail in the text):
        \textbf{(A)} a small portion of a tree sequence without unary nodes;
        \textbf{(B)} the implied inheritance pattern of the two portions of the haplotype carried by ancestral node 4,
        labeled $L$ and $R$;
        \textbf{(C)} marginal trees with a unary node added,
        which produces \textbf{(D)} a more parsimonious haplotype inheritance pattern
        (that also includes fewer edges).
        \label{fig:conceptual}
    }
\end{figure}

@@

% * description (HALLEY)
% * proof of something:
%     - arrives at a local minimum?
%     - guess at typical reduction? (reduces edges by 1/3?)
    We define a \textit{tree sequence}, denoted $\T$, 
    as a tuple $\left(T_1,...,T_{|\T|}}\right)$ with a sequence of points
	$0 = a_0 \leq a_1 \leq \cdots < a_{|\T|} = L$ where
     each $T_k$ is a tree describing genealogical relationships between samples
     over a distinct portion of the genome $[a_{k-1}, a_k)$.
    Given any tree sequence, our goal is to
    identify areas of implied inheritance of haplotypes.
    We do so by identifying short paths (2 edges) in a tree.
    If neighboring trees contain edges with 
    identical ancestor (parent) and descendant (child) nodes, 
    we assume that there is implied coalescence of haplotypes between nodes, 
    and we will extend the path's edges to the neighboring tree \ref{fig:extending_diagram}.
  	In more concrete terms, for a tree $T_k$ in the tree sequence $\T$ 
    suppose there exists a path which contains a node $3$,
    with parent and child nodes $4$ and $0$, respectively.
    If there is an edge between the nodes $4$ and $0$ in trees 
    $T_{k+1}$ (or $T_{k-1}$),
    then we wish to extend the edges $4\to 3$ and $3\to 0$ 
    into tree $T_{k+1}$ and then remove the edge $4 \to 0$ from $T_{k+1}$. 
    This action reduces the length of $4 \to 0$ on the genome,
    and in some cases, completely removes the edge. 
    We now perform this action on all such ancestral paths 
    over the entire tree sequence.
    With this algorithm \ref{alg:edge} we only extend existing edges
    over a larger interval on the genome,
    and remove unnecessary edges in the process.

\begin{figure}[!ht]
\begin{center}
	\includegraphics[width=5in]{edge_extend_method.pdf}
\end{center}
\caption{A visualization of the \textit{extend edges} method. Given two neighboring trees $T_1$ and $T_2$, the tree $T_1$ has a short path of edges $4 \to 3\to0$. The next tree contains a single edge $4 \to 0$. If node $3$ is not in $T_2$, then edge extend will extend edges $ 4\to 3$ and $3\to 0$ to tree $T_2$, hence removing the edge $4\to 0$. 
    \label{fig:extending_diagram}
}
\end{figure}

@@

\caption{Extends edges across a tree sequence $TS$.}\label{alg:edge}
\SetKwFunction{ex}{Extend Edges}
\SetKwFunction{fwd}{Extend}
\DontPrintSemicolon
\LinesNotNumbered
%\SetAlgoNoEnd
\input{A tree sequence $\T$}
\output{New tree sequence,$\T'$}

@@

\Fn{\fwd (tree sequence $\T$, forwards = \tn{True})}{
{	
\eIf{forwards=\tn{True}}{$\T = \left(T_1,...,T_{|\T|}\right)$\BlankLine}
{\BlankLine
	$\T = \left(T_{|\T|},..., T_1\right)$\;}
}
\Indm \For{$T_k\in \T$}{
	Find all edge pairs $(e_1,e_2)$ such that $e_1 = a\to b$ and $e_2 = b\to c$ for some nodes $a,b,c$.\;
	\For{edges $f$ in $T_{k+1}$}{
		\If{$f = a \to c$ and $b$ is not in $T_{k+1}$}{
		Remove $f$ from $T_{k+1}$.\;
		Add $e_1$ and $e_2$ to $T_{k+1}$.\;
		Update $\T$.\;
	}
	}
}\;						
\KwRet{$\T$}\;
}
\BlankLine						
\Fn{\ex ($\T$, iter=100)}{
	\For{$*$ \KwTo \Range(iter)}{
		$\T' \gets \T$.\fwd(fowards=\tn{True})\;
		$\T' \gets \T$.\fwd(fowards=\tn{False})\;
		\If{Number of edges in $\T$ = Number of edges in $\T'$}{
			\Break\BlankLine}
		}
}
\KwRet{$\T'$}\;
\end{algorithm2e}

%% Find counter example to global minimum.

@@

\par Now with an algorithm in place
 we need a metric to discern
 if trees modified with edge extend
 in fact correctly infer haplotypes in our original tree sequence.
 One such function a so called \textit{discrepancy function}, can be called on a pair of tree 
 To count all the discrepancies we must begin by counting differences for each node in the tree sequence.

\begin{figure}

    \caption{
        Conceptual figure displaying what discrepancy is?
        TODO: brainstorm.
        \label{fig:conceptual_discrepancy}
    }
\end{figure}
